---
title: 'Homework: numerical linear algebra and stability issues'
output: html_notebook
---

```{r setup, echo=FALSE}
required_packages <- c('MASS')
for (pkg in required_packages) {
  if (!(pkg %in% rownames(installed.packages()))) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

source(file.path("..", "R", "colors.R"))
```


# Problem 1: Kriging / Gaussian process regression

## Part 1: Finishing unfinished business from lecture

Execute a command `git grep -n "sq_exp_cov <- function"` from the root directory of the git repository.
Paste the output below.

```{zsh, eval=FALSE}
# Paste the `git grep` output.
```

The grep output tells you where you can find the (failed) kriging / Gaussian process regression example in the lecture source file.
Copy the synthetic observation from there (the one defined on the range $[0, 0.4]$ and $[0.6, 1.0]$).

```{r}
# Fill in
y_obs <- mvrnorm(mu = rep(0, n_obs), Sigma = Sigma_obs)
```

Let's now interpolate the missing values in $(0.4, 0.6)$ for real.
To this end, we use eigen decomposition / pricipal component analysis of the GP.
We can ignore the components with negligible ($\approx 0$) variance in computing $\boldsymbol{\Sigma}_{\textrm{obs}}^{-1} \boldsymbol{y}_{\textrm{obs}}$ since those components will not affect the final results of kriging/interpolation.
In other words, we approximate $\boldsymbol{\Sigma}_{\textrm{obs}}^{-1}$ with $\sum_{i = 1}^k \lambda_i^{-1} \boldsymbol{u}_i \boldsymbol{u}_i^\intercal$ in computing $\boldsymbol{\Sigma}_{\textrm{obs}}^{-1} \boldsymbol{y}_{\textrm{obs}}$ for a suitably chosen number of principal components $k$.
Choose $k$ to capture $99.99$\% of variance in $\boldsymbol{y}_{\textrm{obs}} \sim \mathcal{N}(\boldsymbol{0}, \boldsymbol{\Sigma}_{\textrm{obs}})$.

```{r}
loc_new <- seq(.41, .59, .01)
# Compute the conditional mean of GP on the new locations.
```

```{r, fig.dim=c(8, 5), fig.align='center'}
solid_circle_index <- 19
plot(
  # Fill in
  xlab="s", ylab="y(s)", 
  pch=solid_circle_index,
  cex.lab=1.4, cex.axis=1.4, 
  col=jhu_color$heritageBlue, 
  frame.plot = F
)
points(
  # Fil in
  col = jhu_color$spiritBlue
)
```

## Part 2: Quantifying uncertainty in interpolation
Use the eigen-decomposition technique to compute the conditional covariance 
$$\textrm{Var}(\boldsymbol{y}_1 \, | \, \boldsymbol{y}_2)
  = \boldsymbol{\Sigma}_{11} - \boldsymbol{\Sigma}_{12} \boldsymbol{\Sigma}_{22}^{-1} \boldsymbol{\Sigma}_{21},$$
calculate 95% CI intervals of the interpolation, and plot them.

## Part 3: Kriging based on an alternative GP covariance function

Albeit popular in certain fields, GP based on squared exponential covariance function is often considered "too smooth" for many applications.
In particular, its smoothness is partially responsible for the extreme ill-conditioning of its covariance matrix. 
Here we try carrying out the same interpolation task instead using a _Matern_ covariance function. 

First find the definition and implementation of a Matern covariance using a command `git grep -n "Matern"`.

```{r}
# Definition of the `matern_cov` function.
```

Then use the eigen decomposition technique as before for interpolation and uncertainty quantification.
Afterward, check the condition number of $\boldsymbol{\Sigma}_{\textrm{obs}}$ and see if you can directly compute $\boldsymbol{\Sigma}_{\textrm{obs}}^{-1} \boldsymbol{y}_{\textrm{obs}}$ using the `solve` function.
Do we get the same answer as the one based on eigen decomposition?

```{r}
# Fill in
```

**Bonus question:** Symmetric positive-definite matrices can be inverted more quickly via the _Cholesky decomposition_ $\boldsymbol{L} \boldsymbol{L}^\intercal = \boldsymbol{A}$ than via the LU or QR decomposition. 
<!-- The QR and LU decomposition takes $O(2 n^3 / 3)$ while the Cholesky takes $O(n^3 / 3)$ operations. -->
The Cholesky decomposition also provides the "sqrt" of covariance matrix you need to sample from multivariate Gaussians.
(Though you might be stuck with the eigen decomposition technique if the matrix is ill-conditioned.)
Try computing $\boldsymbol{\Sigma}_{\textrm{obs}}^{-1} \boldsymbol{y}_{\textrm{obs}}$ via `chol` and `forward/backsolve`.