---
title: "Performant Statistical Computing and Algoware Development"
subtitle: "where statistical computing, algorithm, and software meets"
author: "Aki Nishimura"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "extra.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      slideNumberFormat: "%current%"
      countIncrementalSlides: false
    includes:
      before_body: mathjax_config.html
---

```{r setup, include = FALSE}
required_packages <- c('Matrix')
for (pkg in required_packages) {
  if (!(pkg %in% rownames(installed.packages()))) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}
source(file.path("..", "R", "colors.R"))
```

# What's wrong with Statistics

How long does matrix multiplication take in base R

```{r, eval=FALSE}
n <- 4096L
A <- matrix(rnorm(n * n), nrow = n, ncol = n)
B <- matrix(rnorm(n * n), nrow = n, ncol = n)
system.time(A %*% B)
```

... as opposed to Anaconda Python?

```{python, eval=FALSE}
import numpy as np
n = 4096
A = np.random.randn(n, n)
B = np.random.randn(n, n)
%timeit A.dot(B)
```

---

# What's wrong with Statistics

```{r}
n_obs <- 5
n_pred <- 3
n_nonzero <- 5
col_index <- c(1, 1, 1, 2, 2, 3)
row_index <- c(1, 3, 5, 2, 4, 1)
sparseMatrix(i = row_index, j = col_index, x = 1)
```

---

# What's wrong with Statistics

```{r}
n_obs <- 10^6
n_pred <- 10^3
density <- .01
n_nonzero <- as.integer(n_obs * n_pred * density)
```

```{r, eval=FALSE}
# Simulate sparse design matrix
row_index <- sample.int(n_obs, n_nonzero, replace = TRUE)
col_index <- sample.int(n_pred, n_nonzero, replace = TRUE)
X <- sparseMatrix(i = row_index, j = col_index, x = 1)
```

--
How much memory to store $X$ in dense format?

```{r}
memory_req_in_gigabytes <- 8 * n_obs * n_pred / 1024^3
sprintf("Dense storage will require %.2gGB.",  memory_req_in_gigabytes)
```

---

# What's wrong with Statistics

```{r, eval=FALSE}
X <- sparseMatrix(i = row_index, j = col_index, x = 1)

# Simulate (completely random) survival time
failure_time <- rexp(n_obs)
censoring_time <- rexp(n_obs)
observed <- rbinom(n_obs, 1, prob = .7)
y <- data.frame(
  time = ifelse(observed, failure_time, censoring_time),
  status = observed
)
y <- as.matrix(y)

glmnet(X, y, family = 'cox') # Older versions throw errors
```

---

# Words from Joel Spolsky, <br> $\qquad$co-founder of Stack Overflow

.center[
_"Nobody would be surprised if you told them it was hard to do open heart surgery. OK, fine, so maybe software isn't open heart surgery. But it has about the same number of moving parts, it's probably just as complicated, and it's critical in a different way._

_But there's something weird about designing software specifically: Everybody thinks they know how to do it, even when they have no training. They just think, Oh sure, I can do that!"_
]

---
class: middle

# So, what's your point, dude?

--

Software is hard (but thanks Hadley for your help).

--

Computing is hard (who's going to be a Hadley here?)

--

Statisticians don't even know what they don't know.

--

Statisticians' needs left out while soft/hardware evolves.

--

We gotta start changing this.

